Abstractions for Multi-Sorted Substitutions

### 2023-08-26 paper/paper.tex
Update.

\usepackage{agda-unicode}

@@

  % The following line will be replaced by make with an \input{} for each
  % .lagda.tex file:

  % SED MARKER

  \AgdaNoSpaceAroundCode{}

@@

  The term type \texttt{S~⊢~s} is then indexed by its sort \texttt{s} and the sorts of its free variables \texttt{S}.
  For example, \texttt{[𝕥, 𝕥] ⊢ 𝕖} is the type of expressions (\texttt{𝕖}) with two free type-variables (\texttt{𝕥}).
  \newpage
  \AppCode\FSyntax
  The notation \texttt{\_⊢\_} is often used for terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \texttt{s} correspond to types, and lists of sorts \texttt{S}
  correspond to type environments.

  As it is typical in intrinsic typing, variables are represented as
  typed (in our case sorted) DeBruijn indices \texttt{S~∋~s}, i.e.\
  values of the usual proof-relevant list-membership relation:
  \LibCode\KVariables

  Contrary to the usual intrinsic typing, we limit the environment
  \texttt{S} to sorts with index \texttt{Var}.
  This ensures that it is impossible to construct kind variables, as
  it is not possible to extend \texttt{S} with the sort of kinds
  \texttt{𝕜}.

  \subsection{A Structure for Multi-Sorted Syntax}
  The multi-sorted syntax makes it easy to capture general syntax definitions
  in a structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \texttt{`\_} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

@@

  A renaming \texttt{S₁ →ᵣ S₂} maps variables from \texttt{S₁} to
  variables from \texttt{S₂}.
  A substitution \texttt{S₁ →ₛ S₂} maps variables from \texttt{S₁} to
  terms with free variables from \texttt{S₂}.

@@

  \subsection{The Need for Renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \texttt{σ} to a term
  \texttt{t} cannot be defined directly by structural recursion on
  \texttt{t}.
  If \texttt{t} is a term like \texttt{λx e},
  where the subterm \texttt{e} is under a binder, then the variables
  in \texttt{e} are shifted, i.e.\ the variable \texttt{zero} refers
  to the variable from the binder, whereas a variable \texttt{suc x}
  refers to the variable \texttt{x} from outside of the binder.
  As the terms contained in \texttt{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \texttt{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \texttt{x} to the term \texttt{` (suc x)}.
  However, applying this shifting substitution to the terms in
  \texttt{σ} is not structurally recursive, as the terms in \texttt{σ}
  are not subterms of \texttt{λx e}.

  The usual workaround is to first define renamings,
  i.e.\ substitutions that replace variables with variables.
  In contrast to general substitutions, renamings can be easily
  shifted: they only contain variables \texttt{x}, which can be
  shifted by simply taking their successor \texttt{suc x}.
  As the shifting substitution is a renaming, we can then define substitution,
  where we apply the shifting \emph{renaming} to the terms in \texttt{σ} when
  the substitution needs to go under a binder.

@@

  % A \texttt{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \texttt{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instanciate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

  As we intend to have exactly two kit instances, we choose names of
  the form \texttt{x/y}, where \texttt{x} is the name we choose for the
  variable instance, and \texttt{y} is the name we choose for the term
  instance.
  For example the parameter type \texttt{\_∋/⊢\_} will be instanciated
  to \ACon{\_∋\_} for the variable kit, and to \AField{\_⊢\_} for the
  term kit.

@@

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \AField{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \ACode{`\_}.
  \item \AField{`/id-injective} and \AField{id/`-injective} state that
    \AField{`/id} and \AField{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \AField{wk-id/`} characterizes the behaviour of the \AField{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \texttt{S₁ →ᵣ S₂} and
    substitutions \texttt{S₁ →ₛ S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \texttt{S₁ →ₖ S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \texttt{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \texttt{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \texttt{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \texttt{wk} on terms:
    \LibCode\KWeaken
  \end{itemize}

@@

  The only task left now is to create the two instances for the kits
  and define how a map can be applied to a term.
  While the variable kit can be defined as expected, the term kit
  raises a dependency problem: The \texttt{wk} operation for terms requires us
  to already know how to apply a renaming to term.
  To solve this problem, McBride came up with a clever solution:
  Applying a map to a term, can be defined without even having any Kit-instances yet.
  Afterwards we can first define the variable kit, and in the
  definition of the term kit, we are then already allowed to apply renamings.
  After the definition of the term-kit we are then allowed to apply
  both renamings and substitutions as both kits defined.

>>>>

\usepackage{agda-unicode}
\usepackage{agda-generated}
\AgdaNoSpaceAroundCode{}

@@

@@

  The term type \ACode{S \ACon{⊢} s} is then indexed by its sort
  \ACode{s} and the sorts of its free variables \ACode{S}.
  For example, \ACon{[𝕥, 𝕥] ⊢ 𝕖} is the type of expressions (\ACon{𝕖}) with two free type-variables (\ACon{𝕥}).
  \newpage
  \AppCode\FSyntax
  The notation \ACon{\_⊢\_} is often used for terms in
  intrinsically-typed languages. This is no accident: in effect, we
  defined an intrinsically-typed language with the twist that the
  typing relation assures exactly that the syntactic categories are followed.
  Sorts \ACode{s} correspond to types, and lists of sorts \ACode{S}
  correspond to type environments.

  As it is typical in intrinsic typing, variables are represented as
  typed (in our case sorted) DeBruijn indices \ACode{S \ACon{∋} s}, i.e.\
  values of the usual proof-relevant list-membership relation:
  \LibCode\KVariables

  Contrary to the usual intrinsic typing, we limit the environment
  \ACode{S} to sorts with index \ACon{Var}.
  This ensures that it is impossible to construct kind variables, as
  it is not possible to extend \ACode{S} with the sort of kinds
  \ACon{𝕜}.

  \subsection{A Structure for Multi-Sorted Syntax}
  The multi-sorted syntax makes it easy to capture general syntax definitions
  in a structure:
  \LibCode\KTerms
  The first three fields record the definitions of sorts, terms, and variable introduction.
  The last field records that variable introduction
  \AField{`\_} is injective, which is trivially true for
  constructors. The instantiation for our System F syntax is
  straightforward:
  \AppCode\FTerms

@@

  A renaming \ACode{S₁ \ADef{→ᵣ} S₂} maps variables from \ACode{S₁} to
  variables from \ACode{S₂}.
  A substitution \ACode{S₁ \ADef{→ₛ} S₂} maps variables from \ACode{S₁} to
  terms with free variables from \ACode{S₂}.

@@

  \subsection{The Need for Renamings}
  A problem with mechanizing languages with DeBruijn-indices is that
  the operation of applying a substitution \ACode{σ} to a term
  \ACode{t} cannot be defined directly by structural recursion on
  \ACode{t}.
  If \ACode{t} is a term like \ACode{\ACon{λx} e},
  where the subterm \ACode{e} is under a binder, then the variables
  in \ACode{e} are shifted, i.e.\ the variable \ACon{zero} refers
  to the variable from the binder, whereas a variable \ACode{\ACon{suc} x}
  refers to the variable \ACode{x} from outside of the binder.
  As the terms contained in \ACode{σ} may themselves contain free variables,
  they need to be shifted, too, before they can be substituted
  for variables in \ACode{e}. 
  Shifting the variables in a term is itself a substitution, which maps
  each variable \ACode{x} to the term \ACode{\ACon{`} (\ACon{suc} x)}.
  However, applying this shifting substitution to the terms in
  \ACode{σ} is not structurally recursive, as the terms in \ACode{σ}
  are not subterms of \ACode{\ACon{λx} e}.

  The usual workaround is to first define renamings,
  i.e.\ substitutions that replace variables with variables.
  In contrast to general substitutions, renamings can be easily
  shifted: they only contain variables \ACode{x}, which can be
  shifted by simply taking their successor \ACode{\ACon{suc} x}.
  As the shifting substitution is a renaming, we can then define substitution,
  where we apply the shifting \emph{renaming} to the terms in \ACode{σ} when
  the substitution needs to go under a binder.

@@

  % A \ACode{Kit} is a structure that captures the commonalities
  % between variables and terms:
  A \ACode{Kit} is a structure that allows to abstract over whether
  something is a term or a variable. The intention is to instanciate
  this structure exactly twice (once for variables and once for terms),
  and then write definitions, which are parameterized over a kit and
  consequently can be used for both variables and terms.
  \LibCode\KKit

  As we intend to have exactly two kit instances, we choose names of
  the form \ACode{x/y}, where \ACode{x} is the name we choose for the
  variable instance, and \ACode{y} is the name we choose for the term
  instance.
  For example the parameter type \ACode{\_∋/⊢\_} will be instanciated
  to \ACon{\_∋\_} for the variable kit, and to \AField{\_⊢\_} for the
  term kit.

@@

  The operations need to satisfy the following properties:
  \begin{itemize}
  \item \AField{`/`-is-`} states that converting a variable first to a ``variable-or-term'' and
    then further to a term is the same as converting it directly to a
    term using the variable constructor \AField{`\_}.
  \item \AField{`/id-injective} and \AField{id/`-injective} state that
    \AField{`/id} and \AField{id/`} are injective. This property
    follows easily, as both the identity function and the variable
    constructor are injective.
  \item \AField{wk-id/`} characterizes the behaviour of the \AField{wk} function.
    Injecting a variable and then shifting it, is the same as injecting a shifted variable.
  \end{itemize}

  Building on the operations and axioms, we derive further operations:\footnote{
    The definitions are included directly in the record module of the
    Kit, so they are implicitly parameterized over a kit.
  }
  \begin{itemize}
  \item 
    We unify the definition of renamings \ACode{S₁ \ADef{→ᵣ} S₂} and
    substitutions \ACode{S₁ \ADef{→ₛ} S₂}, which differ only in whether they
    return variables or a terms:
    \LibCode\KMap
    We call a value of type \ACode{S₁ \ADef{→ₖ} S₂} a \emph{map}.
  \item 
    A map can be applied to a variable:
    \LibCode\KAp
  \item 
    A map can be extended by a new variable or term:
    \LibCode\KExt
  \item 
    A map can be shifted, as the Kit provides the necessary \AField{wk} operation:
    \LibCode\KWkm
  \item 
    A map can be lifted, which is the operation necessary to
    move a map under a binder of sort \ACode{s}:
    \LibCode\KLift
  \item 
    There is an identity map:
    \LibCode\KId
  \item 
    There is a singleton map, which replaces \ACon{zero} and decreases
    all other variables by one:
    \LibCode\KSingle
  \item 
    There is a map which is equivalent to \AField{wk} on terms:
    \LibCode\KWeaken
  \end{itemize}

@@

  The only task left now is to create the two instances for the kits
  and define how a map can be applied to a term.
  While the variable kit can be defined as expected, the term kit
  raises a dependency problem: The \AField{wk} operation for terms requires us
  to already know how to apply a renaming to term.
  To solve this problem, McBride came up with a clever solution:
  Applying a map to a term, can be defined without even having any Kit-instances yet.
  Afterwards we can first define the variable kit, and in the
  definition of the term kit, we are then already allowed to apply renamings.
  After the definition of the term-kit we are then allowed to apply
  both renamings and substitutions as both kits defined.
